% \sweExpl{Författarnas tack}


% \engExpl{It is nice to acknowledge the people that have helped you. It is
%   also necessary to acknowledge any special permissions that you have gotten –
%   for example, getting permission from the copyright owner to reproduce a
%   figure. In this case, you should acknowledge them and this permission here
%   and in the figure’s caption. \\
%   Note: If you do \textbf{not} have the copyright owner’s permission, then you \textbf{cannot} use any copyrighted figures/tables/\ldots . Unless stated otherwise all figures/tables/\ldots are generally copyrighted.
% }
% \sweExpl{I detta kapitel kan du ev nämna något om
%   din bakgrund om det påverkar rapporten på något sätt. Har du t ex inte
%   möjlighet att skriva perfekt svenska för att du är nyanländ till landet kan
%   det vara på sin plats att nämna detta här. OBS, detta får dock inte vara en
%   ursäkt för att lämna in en rapport med undermåligt språk, undermålig grammatik och
%   stavning (t ex får fel som en automatisk stavningskontroll och
%   grammatikkontroll kan upptäcka inte förekomma)\\
% En dualism som måste hanteras i hela rapporten och projektet
% }


I would first like to thank my two supervisors on this thesis. First Fredrik Enoksson from the Unit of Digital Learning at KTH, for helping me conceptualise and build an assistant that could actually work in the digital environment at KTH. Additionally, helping me with getting in touch with several examiners who participated in the study, helping me with the logistics of getting access to the course room, and much more. I will miss our weekly catch-up that we’ve had throughout the project.


Secondly, I would like to thank my supervisor on the research in the thesis, Michael Welle from the Division of Robotics, Perception and Learning. Michael has been a great support helping figure out what research I should do within my project. I appreciate that you’ve kept me on track throughout the project, without your guidance I’m sure I would’ve still been building on those early prototypes.


I would like to thank all the participating teachers and TAs. Without the courses who voluntarily allowed the assistant to be deployed in their course rooms this research could not have been completed. These include Marcus Lithander, Malin Jansson, Dena Hussain, Christian Smith, Richard James Glassey,  Patric Jensfelt. An extra big thanks to Antonio Maffei and Fabio Marco Monetti who not only participated but also helped in constructing the study.


Without the funding provided from KTH Innovation the research in this thesis would not have been possible. The grants they provided me last year for kthGPT helped me with that and this thesis. Thanks to Hannes Eder Öhrström for all advice and setting me up with the funding last year.


I would also like to thank Sebastian Ware for inspiring me to pursue this subject for my Master thesis. I’ve learned a lot about LLMs and RAG. I’m not sure I would’ve pursued this thesis unless you advised me to do so, thanks!


The support I’ve had from my friends through all of my five years at KTH have been paramount. I’m confident that I wouldn't have made it without them. Even though we’ve been on our own with the master thesis, and many courses before that, the daily support and camaraderie have made these years some of the best years in my life.


Lastly, I would like to thank my employer, Stockholm Code Group, which has supported me in pursuing my degree. They hired me 6 months before enrolling at KTH, knowing that I would become a fulltime student, not many firms would do so. I’ve spent countless hours studying in our offices, and that office, along with the people in it, have been just as significant in helping me finish this degree. Thanks for supporting my side-hustle as a full time student. A special thanks to Patrick Zeits, who convinced me to pursue an education in the first place. I’m grateful you showed me \textit{why} higher education was useful.


\acknowlegmentssignature