\chapter{Background}
\label{ch:background}

% \sweExpl{Bakgrund}
% \generalExpl{When you do your literature study, you should have a nearly complete Chapters 1 and 2.\\
% You may also find it convenient to introduce the future work section into your report early – so that you can put things that you think about but decide not to do now into this section.\\
% Note that later you can move things between this future work section and what you have done as you may change your mind about what to do now versus what to put off to future work.
% }
% \generalExpl{What does a reader (another x student -- where x is your study line) need to know to understand your report?
% What have others already done? (This is the “related work”.) Explain what and
% how prior work/prior research will be applied on or used in the degree
% project/work (described in this thesis). Explain why and what is not used in
% the degree project and give valid reasons for rejecting the work/research.}

This chapter provides basic background information about xxx. Additionally, this chapter describes xxx. The chapter also describes related work xxxx.



% \sweExpl{Vilken viktig litteratur och
% (forsknings-)artiklar har du studerat inom området (litteraturstudie)? }

\section{Major background area 1}

% \sweExpl{Viktigt bakgrundsområde 1}

There are xxx characteristics that distinguish yyy from other information and communication technology (ICT) system, as shown in Figure~\ref{fig:lotsofstars}. Table \ref{tab:tablecaracteristics} summarizes these characteristics.


\input{content/figures/0-lotsofstars}

% \sweExpl{Massor av stärnor (Inspirerad av figur x.y på sidan z i [xxx])}


\input{content/tables/0-sometbl}

% \sweExpl{Egenskaper}
% \sweExpl{Beskrivning}

\subsection{Subarea 1.1}

Entangled states are an important part of quantum cryptography, but also relevant in other domains. This concept might be relevant for neutrinos, see for example \cite{kim_small-mass_2016}.

\subsection{Subarea 1.1.2}

Computational methods are increasingly used as a third method of carrying out
scientific investigations. For example, computational experiments were used to
find the amount of wear in a polyethylene liner of a hip prosthesis in \cite{maguire_jr_new_2014}.

\subsection{Subarea 1.1.2}

Using the nearest data center may improve performance, see \cite{bogdanov_nearest_2015}


\subsection{Link layer Encapsulation}

\label{sec:llencap}

\subsection{IP packet headers}

\label{sec:ipheaders}

\subsection{Test for accessibility of formulas}

\section{Major background area 2}

\subsection{Network layer security}

...

\section{Related work area}

% \sweExpl{Relaterade arbeten}

...


\subsection{Major related work 1}

Carrier clouds have been suggested as a way to reduce the delay between the users and the cloud server that is providing them with content. However, there is a question of how to find the available resources in such a carrier cloud. One approach has been to disseminate resource information using an extension to OSPF-TE, see Roozbeh, Sefidcon, and Maguire \cite{roozbeh_resource_2013}.

\subsection{Major related work n}

\subsection{Minor related work 1}

...

\subsection{Minor related work n}

\section{Summary}

% \sweExpl{Det är trevligt om detta kapitel
%   avslutas med en sammanfattning. Till exempel kan du inkludera en tabell som
%   sammanfattar andras idéer och fördelar och nackdelar med varje - så som
%   senare kan du jämföra din lösning till var och en av dessa. Detta kommer
%   också att hjälpa dig att definiera de variabler som du kommer att använda
%   för din utvärdering.}

% \engExpl{It is nice to have this chapter conclude with a summary. For
%   example, you can include a table that summarizes other people's ideas and
%   benefits and drawbacks with each - so as later you can compare your solution
%   to each of them. This will also help you define the variables that you will
%   use for your evaluation.}

\cleardoublepage
